---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# enrollcast

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![name status badge](https://rorylawless.r-universe.dev/badges/:name)](https://rorylawless.r-universe.dev/)
[![R-CMD-check](https://github.com/localopen/enrollcast/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/localopen/enrollcast/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/localopen/enrollcast/graph/badge.svg)](https://app.codecov.io/gh/localopen/enrollcast)
<!-- badges: end -->

`enrollcast` projects school enrollment using the cohort survival / grade
progression ratio method, implemented as a matrix projection. It works at any
level of aggregation (school, district, LEA, city-wide) and any number of
grades, because it operates on a single grade-by-year enrollment series.

## Installation

```{r installation, eval = FALSE}
install.packages("enrollcast")
```

Install the development version from GitHub:

```{r installation-dev, eval = FALSE}
# install.packages("pak")
pak::pak("localopen/enrollcast")
```

## Usage

```{r example-usage}
library(enrollcast)

# Historical grade-level enrollment (long format).
history <- data.frame(
  year = rep(2021:2023, each = 3),
  grade = factor(rep(c("K", "1", "2"), 3), levels = c("K", "1", "2")),
  enrollment = c(100, 90, 80, 110, 95, 88, 120, 99, 91)
)

# 1. Calculate progression ratios.
ratios <- progression_ratios(history, method = "mean")
ratios

# 2. Project forward. The entry grade (K) is supplied exogenously.
base <- history[history$year == 2023, c("grade", "enrollment")]
projection <- project_enrollment(
  base = base,
  ratios = ratios,
  horizon = 3,
  entry = c(125, 130, 128),
  start_year = 2023
)
projection
```

Inspect the underlying projection matrix at any time:

```{r example-matrix}
progression_matrix(ratios)
```

### Multiple aggregation units

`enrollcast` projects one series per call. To project many schools or LEAs,
split and map:

```{r example-multiple}
# fmt: skip
school_history <- data.frame(
  school = rep(c("North", "South"), each = 9),
  year = rep(2021:2023, each = 3, times = 2),
  grade = factor(rep(c("K", "1", "2"), times = 6), levels = c("K", "1", "2")),
  enrollment = c(100, 90, 80, 110, 95, 88, 120, 99, 91, 120, 
                 110, 100, 130, 115, 108, 140, 119, 111)
)

projections <- lapply(
  split(school_history, school_history$school),
  function(df) {
    ratios <- progression_ratios(df)
    base <- df[df$year == max(df$year), c("year", "grade", "enrollment")]
    project_enrollment(base, ratios, horizon = 3, entry = rep(100, 3))
  }
)

projections$North
```
